---
title: "Evaluating Asteroid Characteristics and Predicting Hazard Levels Using NASA's Asteroid Classification Dataset"
authors: "Vishal Bakshi, Pratiksha Dange, Sudanshu Dotel, Qibin Huang"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
---
  
```{r init, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
library(ezids)
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(warning = F, results = "hide", message = F)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```

```{css, include=FALSE}
.columns {display: flex;}
h1 {color: #87CEFA}
h2 {color:blue}
```

## Column Definitions: 
1. Neo Reference ID	: ID for the Asteroid.
2. Name	: Name of the Asteroid. 
3. Absolute Magnitude: It tells about the brightness of the star i.e., Bright star has low magnitude and Dim star has high magnitude. 
4. Est Dia in KM(min) : This column tells us the estimated minimum size of the asteroid in terms of KM.
5. Est Dia in KM(max) : This column tells us the estimated maximum size of the asteroid in terms of KM.
6. Close Approach Date : The date when the asteroid was very close to earth's atmosphere.
7. Epoch Date Close Approach :	
8. Relative Velocity km per sec : The velocity of the asteroid
9. Miss Dist.(Astronomical)	
10. Miss Dist.(lunar)	
11. Miss Dist.(kilometers)	
12. Miss Dist.(miles)	
13. Orbiting Body	: Tells us the orbiting body of the celestial body
14. Orbit ID	
15. Orbit Determination Date	
16. Orbit Uncertainity	Minimum 
17. Orbit Intersection	
18. Jupiter Tisserand Invariant	: The JTI is the planet jupyter force/pull of the celestial bodies surrounding them. The least JTI is on the celestial bodies which is far from the jupiter
19. Epoch Osculation :	
20. Eccentricity : Decides the shape of the orbit, i.e., 0 < E < 1 : Ellispe, E = 1 : Parabola, E > 1 : Hyperbola
21. Semi Major Axis :	
22. Inclination	:
23. Asc Node Longitude :	
24. Orbital Period :	
25. Perihelion Distance :	
26. Perihelion Arg :	
27. Aphelion Dist	:
28. Perihelion Time :	
29. Mean Anomaly :
30. Mean Motion	: 
31. Equinox	: 
32. Hazardous: Target variable, tells us if above variables is affecting the asteroid being hazardous to earth. 


## Importing Hazardous/Not Asteroid Classification 
```{r, include=TRUE, echo=TRUE, results='show'}
data1<- read.csv("/Users/qibinhuang/Downloads/Astroid_Hazard_Classification.csv")
Ast_raw <- data.frame(data1)
head(Ast_raw)

```
```{r}
library(dplyr)
Ast_raw <- Ast_raw %>%
mutate(Hazardous = as.integer(Hazardous))#(1 is True, 0 is False)
head(Ast_raw)


```


```{r, include=TRUE, echo=TRUE, results='show'}
Ast_Subset <- subset(Ast_raw, select = c('Neo.Reference.ID', 'Name', 'Absolute.Magnitude', 'Est.Dia.in.KM.min.', 'Est.Dia.in.KM.max.', 'Relative.Velocity.km.per.sec',  'Orbiting.Body', 'Orbit.Uncertainity', 'Minimum.Orbit.Intersection', 'Jupiter.Tisserand.Invariant', 'Epoch.Osculation', 'Eccentricity', 'Semi.Major.Axis', 'Inclination', 'Asc.Node.Longitude', 'Perihelion.Distance' ))
head(Ast_Subset)

```






