---
title: "Evaluating Asteroid Characteristics and Predicting Hazard Levels Using NASA's Asteroid Classification Dataset"
author: "Vishal Bakshi, Pratiksha Dange, Sudanshu Dotel, Qibin Huang"
date: "`r Sys.Date()`"
output: 
  prettydoc::html_pretty:
    theme: cayman
    toc: yes
---


```{r init, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
library(ezids)
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(warning = F, results = "hide", message = F)
options(scientific=T, digits = 3) 
library(dplyr)
library(kableExtra)
library(ggplot2)

# options(scipen=9, digits = 3)  
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```

```{css, include=FALSE}
.columns {display: flex;}
h1 {color: #87CEFA}
h2 {color:blue}
```

<div class="column-definitions">
  <p><strong>1. Neo Reference ID:</strong> ID for the Asteroid</p>
  <p><strong>2. Name:</strong> Name of the Asteroid</p>
  <p><strong>3. Absolute Magnitude:</strong> Indicates the brightness of the asteroid; lower magnitude means a brighter star, higher magnitude means a dimmer star.</p>
  <p><strong>4. Est Dia in KM (min):</strong> Estimated minimum size of the asteroid in kilometers.</p>
  <p><strong>5. Est Dia in KM (max):</strong> Estimated maximum size of the asteroid in kilometers.</p>
  <p><strong>6. Relative Velocity (km per sec):</strong> The velocity of the asteroid.</p>
  <p><strong>7. Orbit Uncertainty Minimum:</strong> A measure of the uncertainty in the asteroid's orbit.</p>
  <p><strong>8. Jupiter Tisserand Invariant:</strong> The JTI measures the influence of Jupiter's gravitational pull on celestial bodies; lower JTI values are farther from Jupiter.</p>
  <p><strong>9. Eccentricity:</strong> Determines the shape of the orbit; 0 < E < 1 indicates an ellipse, E = 1 is a parabola, and E > 1 is a hyperbola.</p>
  <p><strong>10. Semi Major Axis:</strong> Indicates the size of the asteroid's orbit.</p>
  <p><strong>11. Inclination:</strong> Denotes the tilt of the asteroid's orbit.</p>
  <p><strong>12. Perihelion Distance:</strong> Indicates how close the asteroid gets to the Sun during its orbit.</p>
  <p><strong>13. Aphelion Dist:</strong> Specifies how far the asteroid is from the Sun during its orbit.</p>
  <p><strong>14. Perihelion Time:</strong> The time when the asteroid is closest to the Sun.</p>
  <p><strong>15. Mean Motion:</strong> Provides information about how the asteroid moves in its orbit.</p>
  <p><strong>16. Hazardous:</strong> Target variable indicating whether the asteroid is considered hazardous to Earth; 1 for hazardous, 0 for not hazardous.</p>
</div>

<style>
  .column-definitions {
    font-family: Arial, sans-serif;
    padding: 10px;
    border: 1px solid #ccc;
    background-color: #f7f7f7;
  }
</style>



## Importing Dataset

```{r, include=TRUE, echo=TRUE, results='show'}
current_directory <- getwd()
data_path <- "Datasets/Asteroid_Hazard_Classification.csv"
file_path <- file.path(dirname(current_directory), data_path)
Ast_raw <- data.frame(read.csv(file_path))

kable(summary(Ast_raw), "html") %>%
  kable_paper(full_width = FALSE) %>% 
  scroll_box(width = "100%", height = "400px")
```



## Data Preprocessing

### Step : 1 : Converting the Hazard column from string to integer, and into categorical.

```{r, include=TRUE, echo=TRUE, results='asis'}
Ast_raw <- Ast_raw %>%
mutate(Hazardous = as.integer(Hazardous))#(1 is True, 0 is False)
Ast_raw$Hazardous <- as.factor(Ast_raw$Hazardous)
```


### Step : 2 : Dropping columns which are not required for our analysis:

```{r, include=TRUE, echo=TRUE, results='asis'}
# Load required libraries
library(knitr)
library(kableExtra)

selected_columns <- c('Neo.Reference.ID', 'Name', 'Absolute.Magnitude', 'Est.Dia.in.KM.min.', 'Est.Dia.in.KM.max.', 'Relative.Velocity.km.per.sec', 'Orbit.Uncertainity', 'Jupiter.Tisserand.Invariant', 'Eccentricity', 'Semi.Major.Axis', 'Inclination', 'Perihelion.Distance', 'Aphelion.Dist', 'Mean.Motion', 'Hazardous')

Ast_Subset <- subset(Ast_raw, select = selected_columns)

# Create a nicely formatted table
kable(Ast_Subset, "html") %>%
  kable_paper(full_width = FALSE) %>% 
  scroll_box(width = "100%", height = "400px")

```

### Step : 3 : Outlier Check: 

```{r}
find_outliers <- function(x) {
  q1 <- quantile(x, 0.25)
  q3 <- quantile(x, 0.75)
  iqr <- q3 - q1
  lower_bound <- q1 - 1.5 * iqr
  upper_bound <- q3 + 1.5 * iqr
  return(x[x < lower_bound | x > upper_bound])
}

# Find and remove outliers in the minimum diameter
outliers_min <- find_outliers(Ast_Subset$Est.Dia.in.KM.min.)
data_cleaned_min <- Ast_Subset[!(Ast_Subset$Est.Dia.in.KM.min. %in% outliers_min), ]

# Find and remove outliers in the maximum diameter
outliers_max <- find_outliers(Ast_Subset$Est.Dia.in.KM.max.)
data_cleaned_max <- Ast_Subset[!(Ast_Subset$Est.Dia.in.KM.max. %in% outliers_max), ]

par(mfrow=c(2,2))

hist(Ast_Subset$Est.Dia.in.KM.min., main="Minimum Diameter", xlab="Est Dia in KM (min)", col="blue")
hist(data_cleaned_min$Est.Dia.in.KM.min., main="Minimum Diameter (Cleaned)", xlab="Est Dia in KM (min)", col="blue")

hist(Ast_Subset$Est.Dia.in.KM.max., main="Maximum Diameter", xlab="Est Dia in KM (max)", col="blue")
hist(data_cleaned_max$Est.Dia.in.KM.max., main="Maximum Diameter (Cleaned)", xlab="Est Dia in KM (max)", col="blue")


# Q-Q plot for minimum asteroid diameter
#original data
qqnorm(Ast_Subset$Est.Dia.in.KM.min.)
qqline(Ast_Subset$Est.Dia.in.KM.min.)
#clean data
qqnorm(data_cleaned_min$Est.Dia.in.KM.min.)
qqline(data_cleaned_min$Est.Dia.in.KM.min.)


# Q-Q plot for maximum asteroid diameter

#original data
qqnorm(Ast_Subset$Est.Dia.in.KM.max.)
qqline(Ast_Subset$Est.Dia.in.KM.max.)
#clean data
qqnorm(data_cleaned_max$Est.Dia.in.KM.max.)
qqline(data_cleaned_max$Est.Dia.in.KM.max.)


shapiro.test(data_cleaned_min$Est.Dia.in.KM.min.)
shapiro.test(data_cleaned_max$Est.Dia.in.KM.max.)

#the p-values for both the minimum and maximum asteroid diameters are less than 2e-16, indicating that the data significantly deviates from a normal distribution.

```
### Step : 4 : Checking for Normal Distribution using QQ-Plot

```{r, include=TRUE, echo=TRUE, results='asis'}

# Load the ggplot2 package
library(ggplot2)

kable(summary(Ast_Subset), "html") %>%
  kable_paper(full_width = FALSE) %>% 
  scroll_box(width = "100%", height = "400px")

# Create a Q-Q plot
qqnorm(Ast_Subset$Perihelion.Distance, main = "Q-Q Plot of Perihelion Distance")

# Add a Q-Q line (e.g., line for the normal distribution)
qqline(Ast_Subset$Perihelion.Distance, distribution = qnorm)


# Create a Q-Q plot
qqnorm(Ast_Subset$Aphelion.Dist, main = "Q-Q Plot of Aphelion Distance")

# Add a Q-Q line (e.g., line for the normal distribution )
qqline(Ast_Subset$Aphelion.Dist, distribution = qnorm)
```


## Data Visualization : 

## Histogram plots to check the distribution

### Part 1: Checking Histograms for Distributions

#### Part 1.1 : Absolute Magnitude, Estimated Diameter (Min), Estimated Diameter (Max) 

```{r}
library(ggplot2)
# Load the tidyr package
library(tidyr)

# Create a histogram for 'Absolute.Magnitude', 'Est.Dia.in.KM.min.', and 'Est.Dia.in.KM.max'
selected_columns_1 <- Ast_Subset[c('Absolute.Magnitude', 'Est.Dia.in.KM.min.', 'Est.Dia.in.KM.max.')]

# Reshape the data into long format
Ast_Subset_long_1 <- pivot_longer(selected_columns_1, 
                                cols = everything(),
                                names_to = "Variable", values_to = "Value")

# Create the histogram
ggplot(Ast_Subset_long_1, aes(x = Value, fill = Variable, color = "black")) +
  geom_histogram(binwidth = 1.0, alpha = 0.7, position = "identity", show.legend = FALSE) +
  facet_wrap(~Variable, scales = "free") +
  labs(title = "Histogram of Absolute Magnitude, Est.Dia.in.KM.min., Est.Dia.in.KM.max", x = "Value", y = "Frequency") +
  scale_fill_manual(values = rainbow(length(unique(Ast_Subset_long_1$Variable))) ) +
  scale_color_manual(values = "black") +
  theme_minimal() +
  theme(legend.position = "right")


```

### Part 1.2 : Histograms for Velocity, Orbit Uncertainity, Jupiter Tisserand Invariant

```{r, include=TRUE, echo=TRUE, results='asis'}


# Create a histogram for 'Relative.Velocity.km.per.sec', 'Orbit.Uncertainity', and 'Jupiter.Tisserand.Invariant'
selected_columns_2 <- Ast_Subset[c('Relative.Velocity.km.per.sec', 'Orbit.Uncertainity', 'Jupiter.Tisserand.Invariant')]

# Reshape the data into long format
Ast_Subset_long_2 <- pivot_longer(selected_columns_2, 
                                cols = everything(),
                                names_to = "Variable", values_to = "Value")

# Create the histogram
ggplot(Ast_Subset_long_2, aes(x = Value, fill = Variable, color = "black")) +
  geom_histogram(binwidth = 1.0, alpha = 0.7, position = "identity", show.legend = FALSE) +
  facet_wrap(~Variable, scales = "free") +
  labs(title = "Histogram of Relative.Velocity.km.per.sec, Orbit.Uncertainity, Jupiter.Tisserand.Invariant", x = "Value", y = "Frequency") +
  scale_fill_manual(values = rainbow(length(unique(Ast_Subset_long_2$Variable))) ) +
  scale_color_manual(values = "black") +
  theme_minimal() +
  theme(legend.position = "right")


```

### Part 1.3 : Histograms for Eccentricity, Semi Major Axis, inclination. 

```{r, include=TRUE, echo=TRUE, results='asis'}

# Create a histogram for 'Eccentricity', 'Semi.Major.Axis', and 'Inclination'
selected_columns_3 <- Ast_Subset[c('Eccentricity', 'Semi.Major.Axis', 'Inclination')]

# Reshape the data into long format
Ast_Subset_long_3 <- pivot_longer(selected_columns_3, 
                                cols = everything(),
                                names_to = "Variable", values_to = "Value")

# Create the histogram
ggplot(Ast_Subset_long_3, aes(x = Value, fill = Variable, color = "black")) +
  geom_histogram(binwidth = 0.1, alpha = 0.7, position = "identity", show.legend = FALSE) +
  facet_wrap(~Variable, scales = "free") +
  labs(title = "Histogram of Eccentricity, Semi.Major.Axis, Inclination", x = "Value", y = "Frequency") +
  scale_fill_manual(values = rainbow(length(unique(Ast_Subset_long_3$Variable))) ) +
  scale_color_manual(values = "black") +
  theme_minimal() +
  theme(legend.position = "right")



```

### Part 1.4 : Histogram for Perihelion Distance, Aphelion Distance, Mean Motion 

```{r, include=TRUE, echo=TRUE, results='asis'}
# Create a histogram for 'Perihelion.Distance', 'Aphelion.Dist', 'Mean.Motion'
selected_columns_4 <- Ast_Subset[c('Perihelion.Distance', 'Aphelion.Dist', 'Mean.Motion')]

# Reshape the data into long format
Ast_Subset_long_4 <- pivot_longer(selected_columns_4, 
                                cols = everything(),
                                names_to = "Variable", values_to = "Value")

# Create the histogram
ggplot(Ast_Subset_long_4, aes(x = Value, fill = Variable, color = "black")) +
  geom_histogram(binwidth = 0.1, alpha = 0.7, position = "identity", show.legend = FALSE) +
  facet_wrap(~Variable, scales = "free") +
  labs(title = "Histogram of Perihelion.Distance, Aphelion.Dist, Perihelion.Time, Mean.Motion", x = "Value", y = "Frequency") +
  scale_fill_manual(values = rainbow(length(unique(Ast_Subset_long_4$Variable))) ) +
  scale_color_manual(values = "black") +
  theme_minimal() +
  theme(legend.position = "right")
```

### Part : 1.5 : Checking if the target variable is symmetrical/ Asymmetrical. 
```{r, include=TRUE, echo=TRUE, results='asis'}

# Create a pie chart for a binary categorical variable
counts <- table(Ast_Subset$Hazardous)

# Assign labels to the counts
labels <- c("Not Hazardous", "Hazardous")

# Create the pie chart
pie(counts, labels = labels, main = "Distribution of Asteroids by Hazardous & Non-Hazardous Status")

```
When going to the distributions of the hazardous column. It is clearly visible that the Non-Hazardous Asteroids are more compared to Hazardous.
The data is Asymmetrical. 
Considering the 3000 minimum rows rule, leveling/making data symmetrical is not a viable option right now. 


## Understanding Patterns With Respect To Domain Knowledge
### Part: 1: JTI vs Eccentricity
```{r, include=TRUE, echo=TRUE, results='asis'}
# Load the ggplot2 package
library(ggplot2)


# Create a scatter plot with color based on a categorical variable
ggplot(Ast_Subset, aes(x = Jupiter.Tisserand.Invariant, y = Eccentricity, color = Hazardous)) +
  geom_point() +
  labs(title = "Scatter Plot of JTI vs Eccentricity with respect to Hazardous", x = "JTI", y = "Eccentricity") +
  scale_color_discrete(name = "JTI vs Eccentricity") 

```
### Part 2 : Perihelion vs Eccentricity With Respect To Hazardous

```{r, include=TRUE, echo=TRUE, results='asis'}
library(plotly)


# Create a 3D scatter plot
plot <- ggplot(Ast_Subset, aes(x = Eccentricity, y = Perihelion.Distance, z = Hazardous, color = Hazardous)) +
  geom_point() +
  labs(title = "3D Scatter Plot of Perihelion Distance vs Eccentricity with respect to Hazardous", x = "Eccentricity", y = "Perihelion Distance", z = "Hazardous") +
  scale_color_discrete(name = "Perihelion Distance vs Eccentricity")

# Convert the ggplot plot to a 3D plotly plot
plotly_plot <- ggplotly(plot)

# Show the interactive 3D plot
plotly_plot

```
According to the domain knowledge/math E(Eccentricity) > 0 forms an elliptical shape. 
As the data is Asymmetrical with 70% (Non-Hazardous) and 30% (Hazardous). We can see more scatter points of Non-Hazardous. 
Whereas, considering the Hazardous Asteroids, Increase in Eccentricity is directly proportional to the change in perihelion distance. 
Perihelion Distance (Closest point to the Sun): When the Perihelion Distance is less, it directly enters into the Hazardous category. 



```{r}

p1 <- ggplot(Ast_Subset, aes(x = as.factor(Hazardous), y = `Perihelion.Distance`)) +
  geom_boxplot(fill= "skyblue") +
  labs(title = "Perihelion Distance vs Hazardous", x = "Hazardous", y = "Perihelion Distance")

p2 <- ggplot(Ast_Subset, aes(x = as.factor(Hazardous), y = `Aphelion.Dist`)) +
  geom_boxplot(fill= "skyblue") +
  labs(title = "Aphelion Distance vs Hazardous", x = "Hazardous", y = "Aphelion Distance")

p1
p2


```
To Statistically Prove it: 

### SMART Question : 1 : How do the hazard classifications of asteroids relate to their proximity to Earth in terms of Perihelion and Aphelion distances?

** Null Hypothesis **: There is no significant difference in perihelion distance for hazardous and non-hazardous asteroids.
** Alternate Hypothesis **: There is significant difference in perihelion distance for hazardous and non-hazardous asteroids. 

```{r, include=TRUE, echo=TRUE, results='asis'}
t_test_perihelion <- t.test(`Perihelion.Distance` ~ Hazardous, data = Ast_Subset)
t_test_aphelion <- t.test(`Aphelion.Dist` ~ Hazardous, data = Ast_Subset)

```

```{r}
# Show the results
print(t_test_perihelion)
```

```{r}
# Show the results
print(t_test_aphelion)
```
Observation: 

Perihelion Distance: P-Value < Alpha Value. So, we reject null hypothesis. 
There is an observable trend where hazardous asteroids tend to exhibit smaller perihelion distances when compared to their non-hazardous counterparts.

Aphelion Distance: A subtle yet statistically significant variation emerges, with asteroids possessing higher aphelion distances demonstrating a slightly increased likelihood of being classified as hazardous. Nevertheless, it's important to note that the means of the two groups remain closely aligned. Thus, we cannot definitively attribute higher aphelion distance as the sole determinant of an asteroid's hazardous classification.


```{r, include=TRUE, echo=TRUE, results='asis'}
# Create a 3D scatter plot
plot <- ggplot(Ast_Subset, aes(x = Eccentricity, y = Aphelion.Dist, z = Hazardous, color = Hazardous)) +
  geom_point() +
  labs(title = "3D Scatter Plot of Aphelion Distance vs Eccentricity with respect to Hazardous", x = "Eccentricity", y = "Aphelion Distance", z = "Hazardous") +
  scale_color_discrete(name = "Aphelion Distance vs Eccentricity")

# Convert the ggplot plot to a 3D plotly plot
plotly_plot <- ggplotly(plot)

# Show the interactive 3D plot
plotly_plot


```

Eccentricity is directly proportional to the aphelion distance. Less the Eccentricity, more the circular orbit and less the Aphelion Distance. 
According to the plot, majority of the plots are overlapped, so Aphelion distance alone is not effecting the asteroid to be Hazardous. 


```{r, include=TRUE, echo=TRUE, results='asis'}
library(ggplot2)
library(tidyr)


# List of variables for density plots
variables <- c('Absolute.Magnitude', 
               'Relative.Velocity.km.per.sec', 'Jupiter.Tisserand.Invariant', 
               'Inclination', 'Mean.Motion')

# Create density plots for each variable
for (variable in variables) {
  p <- ggplot(Ast_Subset, aes_string(x = variable, fill = 'Hazardous')) +
    geom_density(alpha = 0.5) +
    labs(title = paste('Density Plot of', variable),
         x = variable,
         y = 'Density') +
    theme_minimal()
  
  print(p)
}
```

**Absolute Magnitude Observation:** Hazardous asteroids tend to have a lower absolute magnitude. Explanation: Absolute magnitude is a measure of an asteroid’s brightness. A lower absolute magnitude indicates a brighter object. Since hazardous asteroids are often closer to Earth, they may appear brighter, resulting in a lower absolute magnitude.

**Relative Velocity km per sec Observation:** Hazardous asteroids have a wider range of relative velocities. Explanation: This suggests that hazardous asteroids can be traveling at a variety of speeds relative to Earth. Higher speeds could increase the potential hazard of an asteroid.

**Jupiter Tisserand Invariant Observation:** The distribution seems quite similar for both hazardous and non-hazardous asteroids. Explanation: The Jupiter Tisserand Invariant is used to differentiate between asteroids and Jupiter-family comets. The similar distribution suggests that this variable alone might not be a strong predictor of an asteroid’s hazardous classification.

**Inclination Observation:** Hazardous asteroids have a wider range of inclinations. Explanation: Inclination is the tilt of an asteroid’s orbit plane. A wider range of inclinations for hazardous asteroids indicates that they can come from various directions relative to Earth’s orbital plane.

**Perihelion Distance Observation:** Hazardous asteroids tend to have a smaller perihelion distance. Explanation: A smaller perihelion distance means the asteroid gets closer to the sun, and potentially closer to Earth’s orbit, increasing the likelihood of being classified as hazardous.

**Aphelion Dist Observation:** Hazardous asteroids also tend to have a smaller aphelion distance. Explanation: A smaller aphelion distance means the asteroid’s orbit is closer to the sun throughout its entire orbit, again potentially bringing it closer to Earth.

**Mean Motion Observation:** The distribution seems quite similar for both hazardous and non-hazardous asteroids. Explanation: Mean motion is the average motion of the asteroid. The similar distribution indicates that mean motion might not be a significant factor in determining whether an asteroid is hazardous.


## SMART QUESTIONS 

### Question : 2 : Does the impact of Perihelion Distance vary with the absolute magnitude.

```{r, include=TRUE, echo=TRUE, results='asis'}

ggplot(Ast_Subset, aes(x = `Perihelion.Distance`, y = `Absolute.Magnitude`, color = Hazardous)) +
  geom_point(alpha = 0.6) +
  labs(title = "Perihelion Distance vs. Absolute Magnitude", x = "Perihelion Distance", y = "Absolute Magnitude") +
  theme_minimal()

```
Perihelion Distance vs. Absolute Magnitude Description: This plot illustrates how the brightness of an asteroid (in terms of absolute magnitude) relates to its perihelion distance, which is the closest point in its orbit to the Sun. Potential Insight: Hazardous asteroids might tend to have smaller perihelion distances, indicating they come closer to the Sun (and potentially Earth’s orbit). Brighter asteroids (with lower absolute magnitude) might be more easily detected, and their orbits more precisely determined, leading to a higher likelihood of being classified as hazardous if they come close to Earth.


### Question : 3 : Is there any significant difference in the orbit uncertainty, Mean Motion, and inclination with a change in Perihelion Distance? <br> - Reason: Here we are examining the relationship between orbit uncertainty, mean motion, and inclination and perihelion distance (also a numeric variable). <br>

```{r , include=TRUE, echo=TRUE, results='show'}

# Perform ANOVA for 'OrbitUncertainty' with 'PerihelionDistance'
orbitU_perihelionDist_anova <- aov(`Orbit.Uncertainity` ~ `Perihelion.Distance`, data = Ast_raw)
summary(orbitU_perihelionDist_anova)

# Perform ANOVA for 'MeanMotion' with 'PerihelionDistance'
meanM_perihelionDist_anova <- aov(`Mean.Motion` ~ `Perihelion.Distance`, data = Ast_raw)
summary(meanM_perihelionDist_anova)

# Perform ANOVA for 'Inclination' with 'PerihelionDistance'
inclination_perihelionDist_anova <- aov(Inclination ~ `Perihelion.Distance`, data = Ast_raw)
summary(inclination_perihelionDist_anova)

```

**Test 1 - Orbit Uncertainty vs. Perihelion Distance:**

-   Df (Degrees of Freedom): There is 1 degree of freedom associated with Perihelion Distance, and 3690 degrees of freedom for residuals.
-   Sum Sq (Sum of Squares): The sum of squares for Perihelion Distance is 381, and for residuals, it's 35756.
-   Mean Sq (Mean Square): The mean square for Perihelion Distance is 381.
-   F value: The F-statistic is 39.4.
-   Pr(\>F): The p-value is approximately 3.9e-10, which is extremely small.

**Interpretation:** In this test, the p-value (Pr(\>F)) is very small (3.9e-10), which indicates that there is a statistically significant difference in "Orbit Uncertainty" with a change in "Perihelion Distance." This suggests that "Perihelion Distance" has a significant impact on "Orbit Uncertainty."

**Test 2 - Mean Motion vs. Perihelion Distance:**

-   Df: 1 degree of freedom for Perihelion Distance and 3690 degrees of freedom for residuals.
-   Sum Sq: The sum of squares for Perihelion Distance is 133, and for residuals, it's 276.
-   Mean Sq: The mean square for Perihelion Distance is 133.
-   F value: The F-statistic is 1777.
-   Pr(\>F): The p-value is less than 2e-16, which is extremely small.

**Interpretation:** In this test, the p-value is also very small (much less than 0.001), indicating a highly significant difference in "Mean Motion" with a change in "Perihelion Distance." Thus, "Perihelion Distance" significantly affects "Mean Motion."

**Test 3 - Inclination vs. Perihelion Distance:**

-   Df: 1 degree of freedom for Perihelion Distance and 3690 degrees of freedom for residuals.
-   Sum Sq: The sum of squares for Perihelion Distance is 1681, and for residuals, it's 447526.
-   Mean Sq: The mean square for Perihelion Distance is 1681.
-   F value: The F-statistic is 13.9.
-   Pr(\>F): The p-value is 2e-04, which is relatively small.

**Interpretation:** In this test, the p-value (Pr(\>F)) is smaller than 0.05, indicating a significant difference in "Inclination" concerning changes in "Perihelion Distance." It implies that "Perihelion Distance" has a statistically significant effect on "Inclination," though the significance is not as strong as in the previous tests.

In summary, based on these ANOVA tests, it appears that "Perihelion Distance" has a significant impact on "Orbit Uncertainty," "Mean Motion," and "Inclination," with "Mean Motion" having the most substantial effect, followed by "Orbit Uncertainty" and then "Inclination." The p-values for all three tests suggest that the relationships are statistically significant.




### Question : 4 : What are the most influential features in predicting whether an asteroid is hazardous ? 
```{r, include=TRUE, echo=TRUE, results='asis', fig.width=10, fig.height=10}

library(corrplot)

# List of specified variables for the correlation plot
specified_variables <- c('Absolute.Magnitude', 'Est.Dia.in.KM.min.', 'Est.Dia.in.KM.max.', 
                         'Relative.Velocity.km.per.sec', 'Minimum.Orbit.Intersection', 
                         'Jupiter.Tisserand.Invariant', 'Eccentricity', 
                         'Semi.Major.Axis', 'Inclination', 'Perihelion.Distance', 
                          'Aphelion.Dist', 'Mean.Motion')

# Calculate the correlation matrix for the specified variables
correlation_matrix_specified <- cor(Ast_raw[, specified_variables], use = "complete.obs")

# Create a correlation plot
corrplot(correlation_matrix_specified, method = "color", tl.cex = 0.7, number.cex = 0.7, addCoef.col = "black")
```

```{r, include=TRUE, echo=TRUE, results='show'}

# Install the following package (if not already)
# install.packages("randomForest")

library(randomForest)

# Splitting the dataset into predictors (X) excluding the target variable
X <- select(Ast_Subset, -c(Hazardous))
Y <- Ast_raw$Hazardous

# Fitting a Random Forest model to determine feature importance
set.seed(123)  # Setting a seed for reproducibility
rf_model <- randomForest(X, Y, ntree = 1000)

# Getting the feature importance
importance <- importance(rf_model)

# Printing importance
print(importance)


library(ggplot2)

# Get feature importance as a data frame
importance_df <- as.data.frame(importance(rf_model))

# Create a custom feature importance plot
ggplot(importance_df, aes(x = reorder(rownames(importance_df), -MeanDecreaseGini), y = MeanDecreaseGini)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(x = "Features", y = "Mean Decrease in Gini Index") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Feature Importance in Asteroid Hazard Prediction")

```

\*\* From the above outputs we can infer the following:\*\*

Absolute Magnitude: This feature holds the highest importance score of 104.07, signifying its significant influence on the model's capability to differentiate between hazardous and non-hazardous asteroids. In other words, the brightness of the asteroid, represented by its absolute magnitude, plays a crucial role in hazard classification.

Est Dia in KM min and Est Dia in KM max: These two features also possess substantial importance scores, 99.01 and 110.43 respectively. This indicates that the estimated diameter of the asteroid in kilometers is of paramount importance when assessing whether an asteroid is hazardous or not.

Relative Velocity km per sec: While still important, this feature has a comparatively lower importance score of 16.29. It suggests that the asteroid's velocity, although not as critical as size and magnitude, does contribute to the model's ability to make predictions regarding hazard levels.

Orbiting Body: This feature's importance score is 0.00, implying that it does not impact the model's hazard classification. It may be a candidate for removal from the feature set since it has no discernible effect on the predictions.

Orbit Uncertainty: With a moderate importance score of 40.64, this feature contributes to determining hazard levels. It suggests that the degree of uncertainty in an asteroid's orbit is a relevant factor in classification.

Minimum Orbit Intersection: This feature carries the highest importance score of 469.03, indicating its central role in classifying asteroids as hazardous or not. The closer an asteroid's orbit intersects with others, the more it influences the hazard classification.

Jupiter Tisserand Invariant, Epoch Osculation, Eccentricity, Semi Major Axis, Inclination, Asc Node Longitude, and Perihelion Distance: These features exhibit varying importance scores. While they may not be as influential as absolute magnitude or asteroid size, they still contribute to the model's predictive ability to some extent. These parameters play a role in shaping the model's understanding of asteroid hazard.








